Using Scipion for Stream Image Processing at Cryo-EM Facilities

### 2018-04-18 2.overall.tex
to home

\subsection{Stream Processing}
The main change introduced in \scipion is \emph{stream processing}, that is, computing on data directly as it is produced or received.
In this way, a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.1, users may find  the first steps in the image processing workflow adapted to streaming, including movie alignment and CTF estimation. We have also implemented particle picking and extraction, that are going to be released soon in the new version 1.2. In the developers version, several 2D classification algorithms are being adapted for this new way of processing.

%\scipion is a multilayered application, the first layer handles the interaction with the user (\emph{presentation layer}), the second saves and retrieves the data used by the application  (\emph{data layer}) and finally, the third layer (\emph{business layer}) process the data following the user instructions.  
\scipion stream implementation make extensive use of threads. %A thread is the smallest unit of processing that can be scheduled by an operating system. 
Using threads, a programmer can effectively create two or more tasks that run at the same time and share the protocol resources (i.e. memory). 
%is based on periodic updates of the protocol input and output data during the image processing. 
%Each protocol execution in \scipion is an independent process that can create processing threads. 
Each protocol that works in streaming mode has one of its threads monitoring input and updating output  data while other threads take care of the data processing. This clear separation between data access and data processing marks an important difference between running a script or using \scipion, since it facilitates a form of computing in which several jobs are executed  concurrently, instead of sequentially. Moreover, \scipion workflow engine allows users to modify the workflow on the fly without modifying a single code line.
%In practice, the latter follows  a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. %Moreover, it also allows to quickly develop new workflows .

@@

\subsection{Metaprotocols}
Since \scipion original publication we have developed many \emph{metaprotocols}, that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

The first type of metaprotocols are called \emph{monitors} and are used to produce live analysis plots, generate reports or raise alerts when some problems are detected. A monitor example is the \protocol{CTF-monitor}, that checks the computed defocus values for each micrograph as they are generated. \protocol{CTF-monitor} may raise an alert if the defocus values are above or below certain thresholds and continuously generate HTML files so that users and staff may easily follow the data acquisition and processing either in-house or remotely. An example of these reports may be seen at \url{http://nolan.cnb.csic.es/scipionbox/} [**ROB NOTE: select a nice one an complete URL, I need to ask to javi chichon]%At the project's end, a review of the processing history is produced. [** URL to report example?]
%This feature was very important since we knew that difference facilities will have different needs and criteria to be fulfilled.

The second set of metaprotocols are grouped under the name \emph{consensus}. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \protocol{consensus picking} will compare the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

@@

\scipion is a flexible framework that allows to create different workflows by choosing among many algorithms at each step. In this work we do not recommend a particular workflow  over others since we realize that different specimens, microscopes or facilities may have different requirements. Nevertheless, since a use case may clarify \scipion capabilities, in the following we describe the pipeline offered by default at the \emph{Centro Nacional de Biotecnología} (CNB) cryoEM Facility.

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently \scipion offers  four  methods for movie alignment: motioncor2 \citep{Zheng2016:motioncor2}, correlation (an unpublished CPU version that implements the global alignment method used by motioncor \citep{Li2015}), unblur \citep{Grant2015:unblur} and optical flow \citep{Abrishami2015:optical-flow}. Then the CTF of the resulting micrographs are calculated. Scipion provides three CTF estimation methods:
CTFFIND4 (Rohou and Grigorieff, 2015), gCTF (Zhang, 2016) and Xmipp CTF estimation (Sorzano et al., 2007; Vargas et al., 2013). At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that reports up to which frequency two estimation of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. %Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

>>>>

\subsection{Stream Processing}
The main change introduced in \scipion is \emph{stream processing}, that is, computing on data directly as it is produced or received.
In this way, a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.2, users may find  the first steps in the image processing workflow adapted to streaming, including movie alignment,  CTF estimation and particle picking and extraction. In the developers version, several 2D classification algorithms are being adapted for this new way of processing.

%\scipion is a multilayered application, the first layer handles the interaction with the user (\emph{presentation layer}), the second saves and retrieves the data used by the application  (\emph{data layer}) and finally, the third layer (\emph{business layer}) process the data following the user instructions.  
\scipion stream implementation make extensive use of threads. %A thread is the smallest unit of processing that can be scheduled by an operating system. 
Using threads, a programmer can effectively create two or more tasks that run at the same time and share the protocol resources (i.e. memory). 
%is based on periodic updates of the protocol input and output data during the image processing. 
%Each protocol execution in \scipion is an independent process that can create processing threads. 
Each protocol that works in streaming mode has one of its threads monitoring input and updating output  data while other threads take care of the data processing. This clear separation between data access and data processing marks an important difference between running a script or using \scipion, since it facilitates a form of computing in which several jobs are executed  concurrently, instead of sequentially. Moreover, \scipion workflow engine allows users to modify the workflow on the fly without modifying a single line of code.
%In practice, the latter follows  a more general and extensible approach that allows users to modify the workflow on the fly without modifying a single code line. %Moreover, it also allows to quickly develop new workflows .

@@

\subsection{Metaprotocols}
Since the \scipion original publication we have developed many \emph{metaprotocols}, that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

The first type of metaprotocols are called \emph{monitors} and are used to produce live analysis plots, generate reports or raise alerts when some problems are detected. A monitor example is the \protocol{CTF-monitor}, that checks the computed defocus values for each micrograph as they are generated. \protocol{CTF-monitor} may raise an alert if the defocus values are above or below certain thresholds and continuously generate HTML files so that users and staff may easily follow the data acquisition and processing either in-house or remotely. An example of these reports may be seen at \url{http://scipion.cnb.csic.es/scipionbox/lastHTMLReport/}% [**ROB NOTE: select a nice one an complete URL, I need to ask to javi chichon]%At the project's end, a review of the processing history is produced. [** URL to report example?]
%This feature was very important since we knew that difference facilities will have different needs and criteria to be fulfilled.

The second set of metaprotocols are grouped under the name \emph{consensus}. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \protocol{consensus picking} will compare, at a given instant, the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

@@

\scipion is a flexible framework that allows to create different workflows by choosing among many algorithms at each step. In this work we do not recommend a particular workflow  over others since we realize that different specimens, microscopes or facilities may have different requirements. Nevertheless, since a use case may clarify \scipion capabilities, in the following we describe the pipeline offered by default at the \emph{Centro Nacional de Biotecnolog\'{i}a} (CNB) cryoEM Facility.

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently \scipion offers  five  methods for movie alignment: motioncorr \cite{Li2015}, motioncor2 \citep{Zheng2016:motioncor2}, correlation (an unpublished CPU version that implements the global alignment method used by motioncor \citep{Li2013:motioncorr}), unblur \citep{Grant2015:unblur} and optical flow \citep{Abrishami2015:optical-flow}. Then the CTF of the resulting micrographs are calculated. Scipion provides three CTF estimation methods:
CTFFIND4 (Rohou and Grigorieff, 2015), gCTF (Zhang, 2016) and Xmipp CTF estimation (Sorzano et al., 2007; Vargas et al., 2013). At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that reports up to which frequency two estimation of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. %Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

### 2018-04-18 3.workflow.tex
to home

\section{\scipion Setup at the Different Facilities}

@@

 %\item \emph{Ume\aa\ Core Facility for Electron Microscopy} (UCEM, \url{http://www.kbc.umu.se/english/ucem})[** Jose Miguel can you verify this?] 
 \item \emph{The Swedish National Cryo-EM Facility}. The Facility has two nodes: at \scilifelab in Stockholm and at Ume\aa\ University (\url{http://www.scilifelab.se}, \url{http://www.kbc.umu.se/english/ucem/cryo-em/})
 \item \emph{ESRF Cryo-Electron Microscope} (\url{http://www.esrf.eu/home/UsersAndScience/Experiments/MX/About_our_beamlines/CM01.html})
 \item \emph{eBIC at Diamond Light Source} ( \url{http://www.diamond.ac.uk/Science/Integrated-facilities/eBIC.html})
 \item \emph{CNB CryoEM Service} (\url{http://www.cnb.csic.es/index.php/en/research/core-facilities/microscopia-crioelectronica})
 \item \emph{Molecular Microscopy Consortium - NIH} (\url{https://www.niehs.nih.gov/research/atniehs/facilities/mmc/index.cfm})[**Pablo can you confirm this?]
 \item \emph{National Cancer Institute - NIH} (\url{https://cmm.nci.nih.gov/})[**Pablo can you confirm this?]

@@

\scipion original design satisfied the requirements of the CNB, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm) where a rigorous booking system is followed. There, \scipion was adapted to interface with this booking platform in order to produce better reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond  where projects are handled by ISPyB (a customized laboratory information management system -LIMS- \citep{Delageniere2011}) and data are saved on a distributed file system. % that heavily penalizes access to lists of files. 
In this environment, \scipion needs to constantly interchange information with ISPyB and with other applications through a message queue system called \emph{ActiveQ}.% and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created.

>>>>

\section{\scipion Setup at Different Facilities}

@@

 %\item \emph{Ume\aa\ Core Facility for Electron Microscopy} (UCEM, \url{http://www.kbc.umu.se/english/ucem})[** Jose Miguel can you verify this?] 
 \item \emph{The Swedish National Cryo-EM Facility}. The Facility has two nodes: at \scilifelab in Stockholm and at Ume\aa\ University (\url{http://www.scilifelab.se}, \url{http://www.kbc.umu.se/english/ucem/cryo-em/})
 \item \emph{ESRF Cryo-Electron Microscope} (\url{http://www.esrf.eu/home/UsersAndScience/Experiments/MX/About_our_beamlines/CM01.html})
 \item \emph{eBIC at Diamond Light Source} ( \url{http://www.diamond.ac.uk/Science/Integrated-facilities/eBIC.html})
 \item \emph{CNB CryoEM Service} (\url{http://www.cnb.csic.es/index.php/en/research/core-facilities/microscopia-crioelectronica})
 \item \emph{Molecular Microscopy Consortium - NIH} (\url{https://www.niehs.nih.gov/research/atniehs/facilities/mmc/index.cfm})
 \item \emph{National Cancer Institute - NIH} (\url{https://www.cancer.gov/research/resources/cryoem})

@@

\scipion original design satisfied the requirements of the CNB, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm) where a rigorous booking system is followed. There, \scipion was adapted to interface with this booking platform in order to produce better reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond  where projects are handled by ISPyB (a customized laboratory information management system -LIMS- \citep{Delageniere2011:ispb}) and data are saved on a distributed file system. % that heavily penalizes access to lists of files. 
In this environment, \scipion needs to constantly interchange information with ISPyB and with other applications through a message queue system called \emph{ActiveQ}.% and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created.

### 2018-04-18 3.workflow_cnb.tex
to home

Although CNB staff try to discourage the use of external USB hard disks versus network as way of retrieving the raw and processing data, this is the preferred option for most users.
USB 3.0 drives can theoretically handle data flows of 300 MB/s, but CNB experience is that, in this busy environment, many of them  have trouble exceeding 50 MB/s. In most cases, 50 MB/s is enough to transfer the data before the following microscope shift starts. Nevertheless, real-world is tough and it happens that two, or even three,  users try to download their data simultaneously. Fortunately, output-NAS works as front-end and isolates the data acquiring and processing infrastructure from the user downloading. Network users access to output-NAS using a chrooted common account. This account only allows rsync connections and make impossible to retrieve data if the name of the project to be downloaded is unknown. In order to speed up transfer weak ciphers (as \emph{arcfour}) has been activated in output-NAS.

@@

\emadmin was born as a script that created a normalized tree of directories  in which to store the movies acquired by the microscope, helped users to choose between a few predetermined image processing workflows and finally launched \scipion. At present it has grown in a client-server application that in addition to the old script capacities it has incorporate the following functions:

\begin{enumerate}
 \item Transparent addition of new workflows. \scipion is able to export any executed workflow as a json file. \emadmin can import this files and use them to create new predetermined workflows.
 \item Automatic creation of pdf user's report containing: microscope acquisition parameters and histograms summarizing data resolution and  defocus. 
 \item The above mentioned information is stored in a database for further analysis. By default the system plots the  average resolution and astigmatism of the different acquisitions versus time
 \item Remote management. \emadmin is a client server application, the client maybe executed in a machine different from the one that executes \scipion
 \item Automatic backup of the \scipion project
 \item \scipion is able to create live html reports on the image acquisition process. \emadmin keeps track of all of them and facilitates the access to this information
\end{enumerate}

>>>>

Although CNB staff try to discourage the use of external USB hard disks versus network as way of retrieving the raw and processing data, this is the preferred option for most users.
USB 3.0 drives can theoretically handle data flows of 300 MB/s, but CNB experience is that, in this busy environment, many of them  have trouble exceeding 50 MB/s. In most cases, 50 MB/s is enough to transfer the data before the following microscope shift starts. Nevertheless, real-world is tough and it happens that two, or even three,  users try to download their data simultaneously. Fortunately, output-NAS works as front-end and isolates the data acquiring and processing infrastructure from the user's downloading activities. Network users access to output-NAS using a chrooted common account. This account only allows rsync connections and make impossible to retrieve data if the name of the project to be downloaded is unknown. In order to speed up transfer weak ciphers (as \emph{arcfour}) has been activated in output-NAS.

@@

\emadmin was born as a script that created a normalized tree of directories  in which to store the movies acquired by the microscope, helping users to choose between a few predetermined image processing workflows and finally launched \scipion. At present it has grown into a web application that in addition to the old script capacities it has incorporate the following functions:

\begin{enumerate}
 \item Transparent addition of new workflows. \scipion is able to export any executed workflow as a json file. \emadmin can import this files and use them to create new predetermined workflows.
 \item Automatic creation of pdf user's report containing: microscope acquisition parameters and histograms summarizing data resolution and  defocus. 
 \item The above mentioned information is stored in a database for further analysis. By default the system plots the  average resolution and astigmatism of the different acquisitions versus time
 \item Remote management. \emadmin is a client server application, the client maybe executed in a machine different from the one that executes \scipion
 \item Automatic backup of the \scipion project
 \item \scipion is able to create live HTML reports on the image acquisition process. \emadmin keeps track of all of them and facilitates the access to this information
\end{enumerate}

### 2018-04-18 3.workflow_diamond.tex
to home

\subsection{Case 3: Scipion at eBIC facility (Diamond Light Source)}

@@

Diamond has thousands of users every year, both academic and commercial, and it is vital that their data is kept secure and confidential. A user management system has been developed which allocates each user with a unique federal identity (IDs), and associates those user IDs with the experiments in which they partake in. Visits are structured such that all raw and processed data collected during an experiment are kept in one directory, and only the associated visitors have read and write permissions in that directory. Users may access their data through ISPyB \citep{ispb2011},  a Laboratory Information Management System (LIMS) combining sample tracking and experiment reporting  %The link between the user and their visits enables data management such as archiving data to a long-term tape storage system, or the Laboratory Information Management System ISPyB (REF. Delageniere et al Bioinformatics 2011). In contrast to the CNB setup, Diamond has integrated the microscopes in much the same fashion to that of the beamlines at the facility, as is shown in figure X

@@

To monitor the progress of the automatic image processing, CNB and \scilifelab cryo-EM services make use of \scipion default monitor system. However, in Diamond, the electronic notebook ISPyB is the standard tool to record and show metadata about the different experiments. In this way, a new monitor has been developed which, through a webservice, allows \scipion to directly interface with ISPyB.  This setup provides the users with a web interface showing general metadata about their experiment as well as  real-time updates on steps of the process and relevant output such as motion corrected images, drift values, CTF images, etc.

>>>>

\subsection{Case 3: Scipion at eBIC (Diamond Light Source)}

@@

Diamond has thousands of users every year, both academic and commercial, and it is vital that their data is kept secure and confidential. A user management system has been developed which allocates each user with a unique federal identity (IDs), and associates those user IDs with the experiments in which they partake in. Visits are structured such that all raw and processed data collected during an experiment are kept in one directory, and only the associated visitors have read and write permissions in that directory. Users may access their data through ISPyB \citep{Delageniere2011:ispb},  a Laboratory Information Management System (LIMS) combining sample tracking and experiment reporting  %The link between the user and their visits enables data management such as archiving data to a long-term tape storage system, or the Laboratory Information Management System ISPyB (REF. Delageniere et al Bioinformatics 2011). In contrast to the CNB setup, Diamond has integrated the microscopes in much the same fashion to that of the beamlines at the facility, as is shown in figure X

@@

To monitor the progress of the automatic image processing, CNB and \scilifelab cryo-EM services make use of \scipion default monitor system. However, in Diamond, the electronic notebook ISPyB is the standard tool to record and show metadata about the different experiments. In this way, a new monitor has been developed which, through a ISPyB python API, allows \scipion to directly interface with ISPyB.  This setup provides the users with a web interface showing general metadata about their experiment as well as  real-time updates on steps of the process and relevant output such as motion corrected images, drift values, CTF images, etc.

### 2018-04-18 3.workflow_scilifelab.tex
to home

\subsection{Case 2: Scipion at The Swedish National Cryo-EM Facility}

@@

At the beginning of 2016, the Cryo-EM facility at \scilifelab became one of the early adopters of \scipion streaming processing. Before, there were some home-made scripts to perform motion correction (with motioncorr) and CTF estimation (with Gctf) but the microscope operators recognized the importance of a more general framework. In particular, they were interested in the possibility of modifying easily the processing workflow as well as provide users and staff with  graphical tools for data analysis and quick feedback regarding the data collection. Initially, the same setup script used at CNB was adopted, but it did not fully satisfy the facility requirements and a new one has been developed. In the following sections we will briefly describe the computational infrastructure at the \scilifelab node and the implementation of a Session Wizard to automate the initial setup while fetching information from other sources external to \scipion. 
%[** thalos falcon III, titan k2 and falcon III]

>>>>

\subsection{Case 2: Scipion at the Swedish National Cryo-EM Facility}

@@

At the beginning of 2016, the Cryo-EM facility at \scilifelab became one of the early adopters of \scipion streaming processing. Before, there were some home-made scripts to perform motion correction (with motioncor2) and CTF estimation (with Gctf) but the microscope operators recognized the importance of a more general framework. In particular, they were interested in the possibility of modifying easily the processing workflow as well as provide users and staff with  graphical tools for data analysis and quick feedback regarding the data collection. Initially, the same setup script used at CNB was adopted, but it did not fully satisfy the facility requirements and a new one has been developed. In the following sections we will briefly describe the computational infrastructure at the \scilifelab node and the implementation of a Session Wizard to automate the initial setup while fetching information from other sources external to \scipion. 
%[** thalos falcon III, titan k2 and falcon III]

### 2018-04-18 4.future.tex
to home

Additionally, in collaboration with the Xmipp team, we are pushing the streaming workflow with the aim to obtain a first initial volume. We are also adding structure modeling capabilities, by integration of some functionality from programs such as Coot \citep{emsley2010:coot}, Refmac \citep{Murshudov1997:refmac}, etc.

>>>>

Additionally, in collaboration with the Xmipp team, we are pushing the streaming workflow with the aim of obtaining a first initial volume. We are also adding structure modeling capabilities, by integrating some functionality from programs such as Coot \citep{emsley2010:coot}, Refmac \citep{Murshudov1997:refmac}, etc.

### 2018-04-18 5.conclusions.tex
to home

 \scipion stream capabilities can handle large and continuously produce volumes of data. The benefits of streaming are clear: (1) it helps to process data as soon as it is generated,  providing almost instant results on data quality; (2) it saves user processing time at their home institutions since part of the processing is done while data is being acquired; and (3) 
 it further reduces the overall experimental time by allowing processing the data while it is transferred to the final user laboratory.

 \scipion has been deployed in cloud and there is a  a public image, with all the required preinstalled cryoEM software,  available at Amazon WebServices EC2 and EGI federated clouds. These images make use of \scipion streaming capabilities, that is, they are able to process data as soon as it has been transfered to an Amazon storage unit, and can be used to handle  production peaks.

>>>>

 \scipion stream capabilities can handle large and continuously produced volumes of data. The benefits of streaming are clear: (1) it helps to process data as soon as it is generated,  providing almost instant results on data quality; (2) it saves user processing time and resources at her home institutions since part of the processing is done while data is being acquired; and (3) 
 it further reduces the overall experimental time by allowing processing the data while it is transferred to the final user laboratory.

 \scipion has been deployed in cloud and there is a public image, with all the required preinstalled cryoEM software,  available at Amazon WebServices EC2 and EGI federated clouds. These images make use of \scipion streaming capabilities, that is, they are able to process data as soon as it has been transfered to an Amazon storage unit, and can be used to handle  production peaks.
